Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Update to Figure 6, adding spatial loss vs binsize to show how density striping is related to loss. Note that this still needs to be fixed since the plot uses spatial loss, not net spatial loss (random loss line wrong)

options(scipen=5, digits=5)
#source in binning functions
source("BinningLossFunctions.R")

@@

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm. This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of distributional scenarios. Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning. Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.

Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform. 100,000 observation pairs were simulated for each data set from the following distributions:

%
<<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data. Density changes are impossible to make out in the solid black areas of the plots. 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=

@@

ggsave("./figure/SimScatters.png", width=8, height=5.5, dpi=300)
@
\begin{figure}[hbtp]
\begin{center}
 \includegraphics[keepaspectratio=TRUE,width=.9\textwidth]{./figure/SimScatters.png}
	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
	\end{center}
\end{figure}
%
%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, hus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.

\ktm{These bivariate distributions were selected for the variety in their defining characteristics. \textit{Scagnostics} - a portmanteau for scatterplot diagnostics - are formalized metrics that have been used to numerically summarize key attributes of scatterplots, including: outying, skewness, clumpiness, sparsity, striation, convexity, skinniness, stringiness and monotonicity (\citealt{wilkinson2005graph};\citealt{wilkinson2008scagnostics}). The scagnostic scores in Table~\ref{tab:scagnostics} were calculated for the simulated data sets using the \texttt{scagnostics} package in R \citep{wilkinson2015package}. They show that these examples explore a variety of skewnesses, convexity and skinniness. This exploration uses bivariately independent distributions, thus the monotonicity and clumpiness are both near zero for these examples. The coarse data is rounded into striatations by definition, however the striation scagnostic score is low due to the relatively small scale of striation versus the magnitude of variability in each distribution. These minor striations will still prove problematic when binning is inconsistent with the data resolution. }

@@

@@

<<UniformStripes, echo=F,include=T,eval=T,fig.width=7, fig.height=3, out.width='.99\\linewidth', fig.pos='hbtp',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Binned scatterplots of coarse uniform data.">>= 
# Commented out code for data generation: run once then recomment
### Create data to demonstrate stripes from improper binwidths
# binoutcoarse1 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="standard")[[1]]
# binoutcoarse4 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="standard")[[1]]
# binoutcoarse5 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="standard")[[1]]
# binoutcoarse1Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="random")[[1]]
# binoutcoarse4Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="random")[[1]]
# binoutcoarse5Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="random")[[1]]
# 
# binoutcoarse1$BinSize <- "1 X 1 Unit" ; binoutcoarse1$BinningType <- "Standard" 
# binoutcoarse4$BinSize <- "4 X 4 Unit" ; binoutcoarse4$BinningType <- "Standard" 
# binoutcoarse5$BinSize <- "5 X 5 Unit" ; binoutcoarse5$BinningType <- "Standard" 
# binoutcoarse1Rand$BinSize <- "1 X 1 Unit" ; binoutcoarse1Rand$BinningType <- "Random" 
# binoutcoarse4Rand$BinSize <- "4 X 4 Unit" ; binoutcoarse4Rand$BinningType <- "Random" 
# binoutcoarse5Rand$BinSize <- "5 X 5 Unit" ; binoutcoarse5Rand$BinningType <- "Random" 
# 
# allbindat <- rbind(binoutcoarse1,binoutcoarse4,binoutcoarse5,
#                    binoutcoarse1Rand,binoutcoarse4Rand,binoutcoarse5Rand)
# write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)
# allbindat <- read.csv("StripeBinData.csv",header=TRUE) 
# NOTE: allbindat saved to BinningLossDemos.Rdata which is read in chunk 1

@@

ggplot(aes(x=binxs, y=binys), data=allbindat) +
  geom_tile(aes(fill=binfreq,height=height, width=width), data=allbindat)+
  facet_grid(BinningType~BinSize, scales="free", space="free")+
  ggtitle(" ") +   xlab("") + ylab("") +
  scale_fill_gradient(low="#56B1F7", high="#132B43",
                      limits=c(0, 450),name="Bin Count") + 
  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
                     aspect.ratio=1,
                     panel.grid= element_blank())

@@

@@

>>>>

options(scipen=5, digits=5)
#source in binning functions
source("BinningLossFunctions.R")
source("BinPackageFunctions.R")

@@

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with that binning algorithm. This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of distributional scenarios. Some choices may be proven optimal through analytical properties, while other are data dependent and require empirical exploration of loss from binning. Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.
%
<<SimDataScatterplots2,echo=F,include=F,eval=T,fig.width=8, fig.height=5, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data. Density changes are impossible to make out in the solid black areas of the plots. 97.4\\%, 98.7\\% and 99.1\\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.">>=

@@

ggsave("./figure/SimScatters.png", width=9, height=5.5, dpi=300)
@
\begin{figure}[hbtp]
\begin{center}
 \includegraphics[keepaspectratio=TRUE,width=.8\textwidth]{./figure/SimScatters.png}
	\caption[Scatterplots of fine and coarse versions of the simulated bivariate data.]{\label{fig:SimDataScatterplots} Scatterplots of fine and coarse versions of the simulated bivariate data. }
	\end{center}
\end{figure}
%
Data sets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform. 100,000 observation pairs were simulated for each data set from the following distributions:

%Density changes are impossible to make out in the solid black areas of the plots. 97.4\%, 98.7\% and 99.1\% of the points are completely hidden behind at least one other point due to over-plotting in the coarse data for the uniform, normal and exponential simulations, respectively.
%
The \textit{data resolution} is defined as the smallest increment between successive data values. To observe loss from binning under more realistic conditions, we create three data sets by rounding the values from the originally simulated data sets to the nearest even number, hus a data resolution of 2 units in each dimension. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution at which data is recorded.

@@

\ktm{These bivariate distributions were selected for the variety in their defining characteristics. \textit{Scagnostics} - a portmanteau for scatterplot diagnostics - are formalized metrics that have been used to numerically summarize key attributes of scatterplots, including: \textit{outying, skewness, clumpiness, sparsity, striation, convexity, skinniness, stringiness} and \textit{monotonicity} (\citealt{wilkinson2005graph};\citealt{wilkinson2008scagnostics}). The scagnostic scores in Table~\ref{tab:scagnostics} were calculated for the simulated data sets using the \texttt{scagnostics} package in R \citep{wilkinson2015package}. They show that these examples explore a variety of \textit{skewnesses}, \textit{convexity} and \textit{skinniness}. This exploration uses bivariately independent distributions, thus the monotonicity and clumpiness are both near zero for these examples. The coarse data is rounded into striatations by definition, however the striation scagnostic score is low due to the relatively small scale of striation versus the magnitude of variability in each distribution. These minor striations will still prove problematic when binning is inconsistent with the data resolution. }

@@

<<UniformStripes, echo=F,include=F, cache=TRUE>>=
# Commented out code for data generation: run once then recomment
### Create data to demonstrate stripes from improper binwidths
# binoutcoarse1 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="standard")[[1]]
# binoutcoarse4 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="standard")[[1]]
# binoutcoarse5 <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="standard")[[1]]
# binoutcoarse1Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,1,1,type="random")[[1]]
# binoutcoarse4Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,4,4,type="random")[[1]]
# binoutcoarse5Rand <- RectBin2d(datlist[[1]]$roundx,datlist[[1]]$roundy,0,0,5,5,type="random")[[1]]
# 
# binoutcoarse1$BinSize <- "1 X 1 Unit" ; binoutcoarse1$BinningType <- "Standard" 
# binoutcoarse4$BinSize <- "4 X 4 Unit" ; binoutcoarse4$BinningType <- "Standard" 
# binoutcoarse5$BinSize <- "5 X 5 Unit" ; binoutcoarse5$BinningType <- "Standard" 
# binoutcoarse1Rand$BinSize <- "1 X 1 Unit" ; binoutcoarse1Rand$BinningType <- "Random" 
# binoutcoarse4Rand$BinSize <- "4 X 4 Unit" ; binoutcoarse4Rand$BinningType <- "Random" 
# binoutcoarse5Rand$BinSize <- "5 X 5 Unit" ; binoutcoarse5Rand$BinningType <- "Random" 
# 
# allbindat <- rbind(binoutcoarse1,binoutcoarse4,binoutcoarse5,
#                    binoutcoarse1Rand,binoutcoarse4Rand,binoutcoarse5Rand)
# write.csv(allbindat,"StripeBinData.csv", row.names=FALSE)
# allbindat <- read.csv("StripeBinData.csv",header=TRUE) 
# NOTE: allbindat saved to BinningLossDemos.Rdata which is read in chunk 1

@@

ggplot(aes(x=binxs, y=binys), data=allbindat) +
  geom_tile(aes(fill=binfreq,height=height, width=width), data=allbindat)+
  facet_grid(BinningType~BinSize, scales="free", space="free")+
  ggtitle(" ") +   xlab("") + ylab("") +
  scale_fill_gradient(low="#56B1F7", high="#132B43",
                      limits=c(0, 450),name="Bin Count") + 
  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
                     aspect.ratio=1,
                     panel.grid= element_blank()) +
  theme(legend.position="bottom")

  ggsave("./figure/CoarseUnif.png", width=6, height=4, dpi=300)

@@

# NEED TO SAVE TO BinningLossDemos.Rdata which is read in chunk 1
losses <- data.frame( bintype=NA, binsize=NA, type=NA, record=NA,
                      avg_loss=NA, avg_standardized_loss=NA,
                      avg_loss_x=NA,avg_loss_y=NA)
counter=0

for(type in "Uniform"){
  for(record in "Coarse"){
    for(binsize in seq(.5,8.5,by=.05)){
      counter = counter + 1
      alldatsub <- alldat[which(alldat$type==type & alldat$record==record),]
      reduced_data <- rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
                                  widthx=binsize,widthy=binsize, output="reduced")
      losses[counter,] <- c("Standard",binsize,type,record, bin_loss(reduced_data))
      counter = counter + 1
      reduced_data <- rand_rect_bin_2d(data=alldatsub, xcol="x", ycol="y",originx=-1,originy=-1,
                                  widthx=binsize,widthy=binsize, output="reduced")
      losses[counter,] <- c("Random",binsize,type,record, bin_loss(reduced_data))
    }
  }
}

ggplot()+ 
  geom_line(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
                color=bintype,group=bintype), data=losses,size=.5) +
  geom_smooth(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
                  color=bintype,group=bintype),size=.25, alpha=.2,
              se=FALSE, data=losses,method="lm")+
  geom_point(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
                 color=bintype,group=bintype),
             size=2,data=losses[losses$binsize %in% c(1,4,5), ])+
#   geom_text(aes(x=as.numeric(binsize), y=as.numeric(avg_loss), 
#                 label=paste(binsize,"X",binsize,sep=""),color=bintype,group=bintype),
#             size=10,data=losses[losses$binsize %in% c(1,4,5), ], 
#             nudge_x=rep())
  theme_bw() +
  scale_color_manual("",values=c("red","black"))+
  scale_x_continuous("Bin Sizes", breaks=1:8, labels=paste(1:8,"X",1:8,sep=""))+
  annotate("text", x=c(3, 6.5), y=c(2.5,1.75), label=c("Random","Standard"),
           color=c("red","black"),size=6)+
  theme(legend.position="none", 
         panel.grid.minor.x=element_blank()) + ylab("Average Spatial Loss")

  ggsave("./figure/CoarseUnifLosses.png", width=4, height=5, dpi=300)

@@

\begin{figure}[hbtp]
\centering
  \subfloat[Binned scatterplots]{\includegraphics[keepaspectratio=TRUE,width=.66\textwidth]{./figure/CoarseUnif.png}}
  \subfloat[Spatial loss vs bin sizes]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/CoarseUnifLosses.png}} 
  \caption[Traditional and adapted scatterplots for games vs. strikeouts data.]{\label{fig:UniformStripes} Binned scatterplots and spatial losses for various sized bins with coarse uniform data. Note the slight dip in loss at even integers, when bins are aligned with data resolution.}}
\end{figure}
